\subsection{Aligning SSU sequences with ssu-align}

We'll use a small dataset to demonstrate how the package works.
The file \prog{seed-15.fa} contains five archaeal sequences, five
bacterial sequences and five eukaryotic sequences from the
SSU-ALIGN v0.1 seed alignments. These seed alignments
were derived from alignments from the \db{CRW} database
\cite{CannoneGutell02} as described in section~\ref{sec:models}.

Pretend that \prog{seed-15.fa} is a set of SSU sequences obtained from
an environmental sampling survey and that we want to analyze. First,
we can run the \prog{ssu-align} program to classify each sequence by
its domain, and create an alignment for each domain by executing the
command:

\user{ssu-align seed-15.fa myseqs}\\

\prog{ssu-align} takes two command line arguments. The first is the
target sequence file. The second is the name of a directory that the
program will create and place its output files into. This directory
should not yet exist\footnote{If the directory does exist,
  \prog{ssu-align} will print an error and exit. To overwrite an
  existing directory, use the \prog{-f} option.}.

The program will first print a header describing the program version
used, command used, current date, and some other information:

\begin{sreoutput}
# ssu-align :: align SSU rRNA sequences
# SSU-ALIGN 0.1 (May 2010)
# Copyright (C) 2010 HHMI Janelia Farm Research Campus
# Freely distributed under the GNU General Public License (GPLv3)
# - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
# command: ssu-align seed-15.fa myseqs
# date:    Wed May  5 13:17:20 2010
#
# Validating input sequence file ... done.
#
# Stage 1: Determining SSU start/end positions and best-matching models...
\end{sreoutput}

In stage 1, the program scans the input sequences with each of the
three default SSU models. This has two purposes.  First, it classifies
each sequence by determining which model in the input CM file is its
``best-matching'' model, defined as
the model that gives the sequence the highest primary sequence-based
alignment score using a profile HMM\@. Secondly, it
defines the start and end points of the SSU sequences based on the
best-matching model's alignment.

Stage 1 takes about 20 seconds on this dataset (on an Intel Xeon 3.0
GHz processor, which I'll use for all of the example runs in this
guide). When it finishes you'll see: 

\begin{sreoutput}
# Stage 1: Determining SSU start/end positions and best-matching models...
#
# output file name         description                                
# -----------------------  -------------------------------------------
  myseqs.tab               locations/scores of hits defined by HMM(s)
  myseqs.archaea.hitlist   list of sequences to align with archaea CM
  myseqs.archaea.fa              5 sequences to align with archaea CM
  myseqs.bacteria.hitlist  list of sequences to align with bacteria CM
  myseqs.bacteria.fa             5 sequences to align with bacteria CM
  myseqs.eukarya.hitlist   list of sequences to align with eukarya CM
  myseqs.eukarya.fa              5 sequences to align with eukarya CM
\end{sreoutput}

This lists and briefly describes the seven output files the program created
in the newly created \prog{myseqs/} subdirectory of the current directory.
The content and format of these files are described
in detail in section~\ref{sec:output}, but I'll briefly describe them 
here. The first file \prog{myseqs.tab} is output from
INFERNAL's \prog{cmsearch} program. The other six files are
model-specific: two files for each model that was the best-matching
model for at least one sequence in the input target sequence file
\prog{seed-15.fa}. The \prog{.hitlist} suffixed files contain a list
of the sequences that match best to the model, and the \prog{.fa}
suffixed files are those actual sequences. If any of the models had
not been the best-matching model to at least one target sequence,
there would be no \prog{.hitlist} or \prog{.fa} files for that
model.

The program will now proceed to stage 2, the alignment stage. This
stage serially progresses through each model that was the
best-matching model for at least one sequence and aligns the
best-matching sequences to that model. The alignments are computed by
scoring a combination of both sequence and secondary structure
conservation (see section~\ref{sec:background} for further
discussion). 
%as opposed to the scoring in stage one which only used sequence
%conservation. 
As the alignment to each model finishes, three new lines
of text, one for each of three newly created files, will appear on the
screen. For this example, alignment to all three models takes about 20
seconds. When it finishes you'll see:

\begin{sreoutput}
# Stage 2: Aligning each sequence to its best-matching model...
#
# output file name         description
# -----------------------  ---------------------------------------
  myseqs.archaea.stk       archaea alignment
  myseqs.archaea.cmalign   archaea cmalign output
  myseqs.archaea.ifile     archaea insert info
  myseqs.bacteria.stk      bacteria alignment
  myseqs.bacteria.cmalign  bacteria cmalign output
  myseqs.bacteria.ifile    bacteria insert info
  myseqs.eukarya.stk       eukarya alignment
  myseqs.eukarya.cmalign   eukarya cmalign output
  myseqs.eukarya.ifile     eukarya insert info
  myseqs.scores            list of CM/HMM scores for each sequence
\end{sreoutput}

The newly created alignments are the \prog{.stk} suffixed files. These
were created by INFERNAL's \prog{cmalign} program. The
\prog{.cmalign} and \prog{.ifile} suffixed files were also output by
\prog{cmalign}. As in stage 1, these files were created in the
\prog{myseqs/} subdirectory of the current directory. We'll go over
the key features of the alignment files next. For more detail on all
other types of output files, see section~\ref{sec:output}.

\subsubsection{Description of alignments}
\label{sec:tutorial-stk}

The alignment files are the most important important type of output
file. Take a look at the archaeal alignment we just created in
\prog{myseqs/myseqs.archaea.stk}. The alignment includes consensus
secondary structure annotation and is in \emph{Stockholm format}.
Stockholm format, the native alignment format used by the HMMER
and INFERNAL packages and by the \db{pfam} and \db{rfam}
databases. The key features of the alignment file are:

\begin{itemize}

%\item The alignment is in an interleaved format, like other
%  common alignment file formats such as CLUSTALW.
%  Lines consist of a name, followed by an aligned sequence;
%  the alignment is split into blocks separated by blank lines.

\item Each sequence occurs on a single line and each line consists of
  the sequence name followed by the aligned sequence\footnote{Alternatively,
  interleaved alignment files can be created using the \prog{-i}
  option. See the \prog{ssu-align} manual page at the 
  end of this guide for more information.}.

\item Gaps are indicated by the characters \prog{.}, or
  \prog{-}.  The \prog{.} characters occur in insert columns,
  while \prog{-} characters occur in consensus
  columns. Page~\pageref{sec:background-columns} in
  section~\ref{sec:background-columns} explains the difference between
  these two types of columns. In short, insert columns are not well
  conserved and typically are gaps in most sequences while consensus
  columns are typically non-gaps in most sequences.
  However, many SSU alignments will have large regions of 100\% gaps
  in both insert and consensus columns at the
  beginning and ends of the alignment.  This will happen if the
  sequences are partial SSU sequences, such as those obtained with PCR
  primers that target well conserved regions within the SSU molecule.

\item Special lines starting with {\small\verb+#=GR+} followed by a
  sequence name and then {\small\verb+PP+} contain posterior
  probabilities for each aligned nucleotide for the sequence they
  correspond to.  These are confidence estimates in the correctness of
  the alignment.  Page~\pageref{sec:background-pp} in
  section~\ref{sec:background} introduces posterior probabilities with
  an example in figure~\ref{fig:ambiguity}.
  Characters in \prog{PP} rows have 12 possible values: \prog{0-9}, \prog{*}, or
  \prog{.} If \prog{.}, the position corresponds to a gap in the sequence. A
  value of \prog{0} indicates a posterior probability of between 0.0 and
  0.05, \prog{1} indicates between 0.05 and 0.15, \prog{2} indicates between
  0.15 and 0.25 and so on up to \prog{9} which indicates between 0.85 and
  0.95. A value of \prog{*} indicates a posterior probability of between
  0.95 and 1.0. Higher posterior probabilities correspond to greater
  confidence that the aligned nucleotide belongs where it appears in
  the alignment.  These confidence estimates can be used to mask the
  alignment to remove columns with significant fractions of
  ambiguously aligned nucleotides as demonstrated below.

\item A special line starting with \prog{\#=GC SS\_cons}
  indicates the secondary structure consensus. Gap characters annotate
  unpaired (single-stranded) columns. Basepairs are indicated by any
  of the following pairs: \prog{<>}, \prog{()}, \prog{[]}, or
  \prog{[]}.

\item A special ``RF'' line starting with {\prog{\#=GC RF}}
  indicates the consensus, or ReFerence, model. Gaps in the RF line
  are \emph{insert} columns, where at least 1 sequence has at least 1
  inserted nucleotide between two consensus positions. Uppercase nucleotides
  in the RF line are well conserved positions in the model; lowercase
  nucleotides are less well conserved.
\end{itemize}

